GPU-Beschleunigung der Material Point Method

### 2018-12-04 mscthesis.tex
metric,optim

Generally these methods are augmented by splitting the grid in 2D/3D blocks making use of the shared memory structure of the GPU \cite{PIC:GPU} \cite{honig2010generic} \cite{MPM:GPU}, often called domain decomposition. Shared memory on GPUs typically ranges from 48-64KB.  The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a block. Since the number of particles per grid point in the MPM-method is generally between 4-10 only very limited block sizes are available \cite{MPM:SNOW}.

@@

\subsection{Newton's Method}\label{sec:newton}

@@

\section{Implementation}\label{sec:implementation}
\subsection{Optimization opportunities}

@@

Register > Shared Memory > Local, Global. Therefore one should prefer register memory for intra-thread caching. Shared-memory should be used for extra caching memory(to reduce register spilling) or inter-thread communication within a block. Global memory should be mostly used at the start and end of a shader.

\textbf{Register spilling:} It follows that one should avoid local memory. Local memory is created due to register spilling and stored off chip like global memory, though it is L2-cached. One can calculate the available number of registers per thread with the CUDA Occupancy calculator \cite{NVIDIA:OCCUPANCY}. If the shader uses more registers the register spills.

@@

\textbf{Memory coalescing:} Between a block of threads and a single thread another important entity exists. NVIDIA GPUs group 32 threads into a synchronous warp, respectively AMD groups 64 threads as a wavefront. Memory operations are issued per warp/wavefront. A memory access uses the L2 caching behavior. A L2-cache-line is 32 Bytes long. Thus eight (u)int/floats, two single-precision vectors or a single double-precision vector fit into one cache-line.

\noindent\begin{minipage}{0.45\textwidth}
\begin{lstlisting}[caption={AoS Layout},language={GLSL},style={GL}]
struct Particle {
  vec4 position;
  vec4 velocity;
} Particles[n];
\end{lstlisting}
\end{minipage}
\hfill\noindent\begin{minipage}{0.45\textwidth}
\begin{lstlisting}[caption={SoA Layout},label={llst:soa},language={GLSL},style={GL}]
struct Container{
  vec4 positions[n];
  vec4 velocities[n];
} Particles;
\end{lstlisting}
\end{minipage}

@@

    \texttt{\#define binary\_op(x,y) x+y}
\end{enumerate}
For debugging we output the shader with all processed includes and let the gcc-preprocessor assemble the shader with options: \texttt{gcc -E -x c -P}. Except for the ommission of the glsl version or extension macros, we did not find any problems in this process.

\textbf{Shared Memory Bank conflicts:}

\textbf{Multiple elements per thread and block size:}

\cite{NVIDIA:BEST:PRACTICE} \cite{AMD:GPU_OPEN}
decorator pattern soa vs aos\\
magic\_get reflection\\
single + double\\
GL\_EXT\_TIMER + NVIDIA Nsight\\
\subsection{Parallel Reduction}
\cite{NVIDIA:PARALLEL_REDUCTION}
\subsection{Counting sort}
Map (Reset) \\
\subsubsection{Bin}
Binning\cite{NVIDIA:BINNING} (coalesced, uncoalesced)\\
\subsubsection{Scan}
Scan\cite{NVIDIA:SCAN}\\
Scan(Raking) \cite{NVIDIA:SCAN_MODERN}
\subsubsection{Reordering}
Coalesced, Uncoalesced
\subsection{MPM-Transfers}
Conjugate gradient method
\subsubsection{Particle to grid transfers}
\subsubsection{Grid to particle transfers}
\subsection{Verifying Simulation}
\begin{itemize}
  \item Conservation equations
    \cite{MPM:APIC}
  \item Singular Value Decomposition
  \item Conjugate gradient
\end{itemize}
\section{Evaluation}
\cite{NVIDIA:PEAK_PERFORMANCE}
\cite{KHRONOS:TIMER_QUERY}
\section{Conclusion}
\subsection{Review}
\subsection{Future Work}
Particle Activation: \cite{MPM:GPU}

>>>>

Generally these methods are augmented by splitting the grid in 2D/3D blocks making use of the shared memory structure of the GPU \cite{PIC:GPU} \cite{honig2010generic} \cite{MPM:GPU}, often called domain decomposition. Shared memory is however limited per GPU (GTX 970, 48KB per block).The grid nodes reached by the support of the interpolation function on each side also have to be loaded into memory, typically called the halo of a block. Since the number of particles per grid point in the MPM-method is generally between 4-10 only very limited block sizes are available \cite{MPM:SNOW}.

@@

\subsubsection{Newton's Method}\label{sec:newton}

@@

\subsection{General-purpose computing on graphics processing units}
\subsubsection{Optimization Opportunies}

@@

Register > Shared Memory > Local, Global.

Therefore one should prefer register memory for intra-thread caching. Shared-memory should be used for extra caching memory(to reduce register spilling) or inter-thread communication within a block. Global memory should be mostly used at the start and end of a shader.

\textbf{Memory coalescing:} Between a single thread and a block of threads another important entity exists. NVIDIA GPUs group 32 threads into a synchronous warp, respectively AMD groups 64 threads as a wavefront. (Memory) operations are issued per warp/wavefront. A memory access uses the L2 caching behavior. A L2-cache-line is 32 Bytes long. Thus eight (u)int/floats, two single-precision vectors or a single double-precision vector fit into one cache-line.

\textbf{Shared Memory bank conflicts:} Shared memory is split into 32 banks of 4 bytes. The hardware groups together unique memory accesses on these banks. E.g. if two threads within a warp access bank 17 a bank conflict occurs and the access needs to be split into two groups. The one exception is a broadcast: all threads within a warp access the same bank. The number of banks is equal to the warp size, s.t. throughput is maximized if every thread does a unique bank access. \cite{NVIDIA:BEST:PRACTICE}

\textbf{Register spilling:} One should avoid local memory. Local memory is created due to register spilling and stored off chip like global memory, though it is L2-cached. One can calculate the available number of registers per thread with the CUDA Occupancy calculator \cite{NVIDIA:OCCUPANCY}. If the occupancy falls off to 0\% register spilling occurs. The occupancy is the amount of active warps divided by the possible maximum amount of warps on a Streaming Multiprocessor. It allows the SM to schedule between warps: If the current warp can not continue (memory latency, barrier etc.) it will switch to another. Occupancy is mostly affected due to resource limits on an SM (shared memory, register) as can be seen in the occupancy calculator. Non-optimal occupancy might not have any performance impact at all. \cite{NVIDIA:ACHIEVED_OCCUPANCY}

@@

\textbf{Block size:} In general the block size should be chosen as small as possible to not hit resource limits. Though we in general had better experiences starting with a block size of 64 as can be seen in the Occupancy calculator. Increasing the block size is useful in combination with shared memory usage. Shared memory access is typically 100x faster than uncached global memory accesses. The more threads of a block can reuse shared memory data the faster the memory throughput will be. Therefore operators which involve filtering (also MPM transfers) can use this behavior.

\textbf{Multiple elements per thread:} Merging threads and simultaneously reducing block size can reduce register pressure if the algorithm allows it. Otherwise merging sequential and parallel work (in shared memory) can improve performance heavily for instruction-limited shaders. Since the block size is limited by the GPU large amounts of data need to be split in tree structures to process in parallel. One can not avoid an indirection of writing-back and loading again from global memory. Therefore algorithms often can improve much further from more sequential work to avoid this indirection.
\begin{flushright}\cite{AMD:GPU_OPEN}\cite{NVIDIA:BEST:PRACTICE}\end{flushright}
\subsubsection{GPU Metrics}
The GPU metric topping all others is the raw (averaged) time spend on a shader: $\Delta t_c$. The following factors matter to set $\Delta t_c$ into perspective:
\begin{enumerate}
  \item For the GPU most interesting is the scalability with increasing input data sizes.
  \item Numerical algorithms may spend additional time computing (implicit) quantities to allow for larger physical time steps: $\Delta t_p$. Therefore one might be interested how long the GPU needs to calculate one physical second:
    \begin{equation}
    \Delta t^c_{1s} = \frac{\Delta t_c}{\Delta t_p}.
    \end{equation}
    Otherwise one would calculate the speedup of procedure 1 over procedure 2 for a direct comparision:

    \begin{equation}
      \text{Speedup of Procedure 1} = \frac{\Delta t_{c1}}{\Delta t_{p1}} \frac{\Delta t_{p2}}{\Delta t_{c2}}.
    \end{equation}
\end{enumerate}
  \begin{tabular}{ | l | l | l | l | l |l|}
    \hline
    $\Delta t_c$ & VRAM SOL\% & SM SOL\%  & L2 SOL\%& Tex SOL\% &Lifetime   \\
    \hline
  \end{tabular}
\clearpage
\section{Implementation}\label{sec:implementation}
\noindent\begin{minipage}{0.45\textwidth}
\begin{lstlisting}[caption={AoS Layout},language={GLSL},style={GL}]
struct Particle {
  vec4 position;
  vec4 velocity;
} Particles[n];
\end{lstlisting}
\end{minipage}
\hfill\noindent\begin{minipage}{0.45\textwidth}
\begin{lstlisting}[caption={SoA Layout},label={llst:soa},language={GLSL},style={GL}]
struct Container{
  vec4 positions[n];
  vec4 velocities[n];
} Particles;
\end{lstlisting}
\end{minipage}

@@

    \texttt{\#define binary\_op(x,y) x+y}
\end{enumerate}
For debugging we output the shader with all processed includes and let the gcc-preprocessor assemble the shader with options: \texttt{gcc -E -x c -P}. Except for the omission of the glsl version or extension macros, we did not find any problems in this process.

decorator pattern soa vs aos\\
magic\_get reflection\\
single + double\\
GL\_EXT\_TIMER + NVIDIA Nsight\\
\subsection{Parallel Reduction}
\cite{NVIDIA:PARALLEL_REDUCTION}
\subsection{Counting sort}
Map (Reset) \\
\subsubsection{Bin}
Binning\cite{NVIDIA:BINNING} (coalesced, uncoalesced)\\
\subsubsection{Scan}
Scan\cite{NVIDIA:SCAN}\\
Scan(Raking) \cite{NVIDIA:SCAN_MODERN}
\subsubsection{Reordering}
Coalesced, Uncoalesced
\subsection{MPM-Transfers}
Conjugate gradient method
\subsubsection{Particle to grid transfers}
\subsubsection{Grid to particle transfers}
\subsection{Verifying Simulation}
\begin{itemize}
  \item Conservation equations
    \cite{MPM:APIC}
  \item Singular Value Decomposition
  \item Conjugate gradient
\end{itemize}
\section{Evaluation}
\cite{NVIDIA:PEAK_PERFORMANCE}
\cite{KHRONOS:TIMER_QUERY}
\section{Conclusion}
\subsection{Review}
\subsection{Future Work}
Particle Activation: \cite{MPM:GPU}
